Basic concepts and tools for the Toki Pona minimal and constructed language: description of the language and main issues; analysis of the vocabulary; text synthesis and syntax highlighting; Wordnet synsets

### 2018-07-05 pack/prv/opt/tokipona/aux/article/acm/sample-acmsmall.tex
passado artigo p template correto p tallip (acmsmall)

\documentclass[acmlarge]{acmart}
\usepackage[utf8]{inputenc}

\usepackage{booktabs} % For formal tables
\newcommand{\tttt}[1]{\texttt{#1}}
\usepackage{subcaption}
% \usepackage{tipa}

@@

%\acmBadgeR[http://ctuning.org/ae/ppopp2016.html]{ae-logo}
%\acmBadgeL[http://ctuning.org/ae/ppopp2016.html]{ae-logo}

% Copyright
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

@@

% Document starts
\begin{document}
% Title portion
\title{Basic concepts and tools for the Toki Pona minimalist and
constructed language}
\titlenote{Description of the language and main issues;
Wordnet synsets; analysis of the vocabulary; synthesis and syntax highlighting of texts}

@@

\begin{abstract}
A minimalist constructed language (conlang)
is useful for experiments and comfortable for making tools.
The Toki Pona (TP) conlang is minimalist both in the vocabulary
(with only 14 letters and 124 words)
and in the (about) 10 syntax rules.
The language is useful for being a used and somewhat established
minimalist conlang with at least hundreds of fluent speakers.
This article exposes current concepts and resources
for TP,
and makes available Python scripted routines for
the analysis of the language,
the synthesis of texts, the specification of syntax highlighting
schemes, and the achievement of a preliminary TP
Wordnet.
Focus is on the analysis of the basic vocabulary,
as corpus analyses were found.
The synthesis is based on sentence templates,
relates to context by keeping track of used words,
and renders larger texts by using a fixed number of phonemes (e.g. for poems)
and number of sentences, words and letters (e.g. for paragraphs).
Syntax highlighting 
reflects morphosyntactic classes given in the official dictionary
and different solutions are described and implemented
in the well-established Vim text editor.
The tentative TP Wordnet is made available in three patterns
  of relations between synsets and word lemmas.
In summary, this text holds potentially novel conceptualizations about,
and tools and results in analyzing, synthesizing and syntax highlighting the
TP language.
\end{abstract}

@@

% DO NOT use this command unless you want to change
% the default behavior
% \authorsaddresses{Authors' addresses: G.~Zhou, Computer Science
%   Department, College of William and Mary, 104 Jameson Rd,
%   Williamsburg, PA 23185, US, \path{gzhou@wm.edu}; V.~B\'eranger,
%   Inria Paris-Rocquencourt, Rocquencourt, France; A.~Patel, Rajiv
%   Gandhi University, Rono-Hills, Doimukh, Arunachal Pradesh, India;
%   H.~Chan, Tsinghua University, 30 Shuangqing Rd, Haidian Qu, Beijing
%   Shi, China; T.~Yan, Eaton Innovation Center, Prague, Czech Republic;
%   T.~He, C.~Huang, J.~A.~Stankovic University of Virginia, School of
%   Engineering Charlottesville, VA 22903, USA; T. F. Abdelzaher,
%   (Current address) NASA Ames Research Center, Moffett Field,
%   California 94035.}

>>>>

\documentclass[format=acmsmall, review=false, screen=true]{acmart}
\usepackage[utf8]{inputenc}

\usepackage{booktabs} % For formal tables
\newcommand{\tttt}[1]{\texttt{#1}}
\usepackage{subcaption}
\usepackage{tipa}

@@

% Copyright
%\setcopyright{acmcopyright}
\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

@@

% Document starts
\begin{document}
% Title portion. Note the short title for running heads
\title[A Multifrequency MAC for Wireless Sensor]{Basic concepts and tools for the Toki Pona minimal and constructed language: description of the language and main issues; analysis of the vocabulary; text synthesis and syntax highlighting; Wordnet synsets}

@@

\begin{abstract}
A minimal constructed language (conlang)
is useful for experiments and comfortable for making tools.
The Toki Pona (TP) conlang is minimal both in the vocabulary
(with only 14 letters and 124 lemmas)
and in the (about) 10 syntax rules.
The language is useful for being a used and somewhat established
minimal conlang with at least hundreds of fluent speakers.
This article exposes current concepts and resources
for TP,
  and makes available Python (and Vim) scripted routines for
the analysis of the language,
synthesis of texts, syntax highlighting schemes,
  and the achievement of a preliminary TP Wordnet.
Focus is on the analysis of the basic vocabulary,
as corpus analyses were found.
The synthesis is based on sentence templates,
relates to context by keeping track of used words,
and renders larger texts by using a fixed number of phonemes (e.g. for poems)
and number of sentences, words and letters (e.g. for paragraphs).
Syntax highlighting 
reflects morphosyntactic classes given in the official dictionary
and different solutions are described and implemented
in the well-established Vim text editor.
The tentative TP Wordnet is made available in three patterns
  of relations between synsets and word lemmas.
In summary, this text holds potentially novel conceptualizations about,
and tools and results in analyzing, synthesizing and syntax highlighting the
TP language.
\end{abstract}

@@

### 2018-07-05 pack/prv/opt/tokipona/aux/article/acm/samplebody-journals.tex
finalizado artigo para publicacao
passado artigo p template correto p tallip (acmsmall)

\section{Introduction}\label{intro}
Toki Pona (TP) is a minimalist conlang (constructed language)
with only 124 words (120 without the synonyms).
Therefore, the concepts are usually very general and different,
and, without context, the words are not frequently related through
meronymy and hyponymy.
Such a linguistic setting is desired because of the simplicity
which entails facilitated e.g. learning and tool making.
Another reason why the minimalist language design is compelling
is the study and harnessing of the strong and weak forms of
the Sapir-Whorf hypothesis (linguistic relativity),
i.e. that language dictates or at least influences thought
and world experience~\cite{sapWho}.
Accordingly, one uses a conlang as a thinking tool (or platform)
or to make experiments about the influence language has on
the thoughts of the `speaker' (also writer and reader).
Toki Pona is often described as a tool to meditate,
to simplify the thinking processes, and as a way to modify
the mood and impressions about the world.~\cite{tpLang,kama,tp3,tp4}
In~\cite{interview}, Sonja Lang (the creator of TP),
describes that she has seen the language been used successfully
in the context of management, creation of texts, legal texts,
etc.

@@

There are standard guidelines for pronunciation,
but the language allows for considerable allophonic
variation.
For example, /p t k s l/ might be pronounced
[p t k s l] or [b d g z !R].
% [p t k s l] or [b d g z \textipa{\!R}].% ſ].
Especially for poetry, one might consider
j and w to be vowels
(e.g. j as 'i' and w as 'u').

@@

\section{Conclusions and further work}\label{conc}
This document presented a potentially novel description of Toki Pona (TP)
language in Section~\ref{basics}, and innovative and useful software routines for dealing with
the language in Section~\ref{hacks}.
As a minimalist conlang, TP is very convenient for cognitive
experiments
through linguistic semantics and
for devising tools for analysis, creation and visualization,
a fact to which this document is evidence.
Other conlangs, and even non-planned languages might benefit from
TP and this content in many ways, e.g. one might
synthesize TP texts and translate them as convenient,
adapt the routines to obtain statistics of the vocabulary,
or take advantage of the language description to devise new conlangs
or stylistic guidelines for any language.

@@

  \item Explore a minimalistic conlang to describe software

@@

% Bibliography
\bibliographystyle{ACM-Reference-Format}
\bibliographystyle{unsrt}
\bibliography{sample-bibliography}

>>>>

\section{Introduction}\label{intro}
Toki Pona (TP) is a minimal conlang (constructed language)
with only 124 words (120 without the synonyms).
Therefore, the concepts are usually very general and different,
and, without context, the words are not frequently related through
meronymy and hyponymy.
Such a linguistic setting is desired because of the simplicity
which entails facilitated e.g. learning and tool making.
Another reason why the minimal language design is compelling
is the study and harnessing of the strong and weak forms of
the Sapir-Whorf hypothesis (linguistic relativity),
i.e. that language dictates or at least influences thought
and world experience~\cite{sapWho}.
Accordingly, one uses a conlang as a thinking tool (or platform)
or to make experiments about the influence language has on
the thoughts of the `speaker' (also writer and reader).
Toki Pona is often described as a tool to meditate,
to simplify the thinking processes, and as a way to modify
the mood and impressions about the world.~\cite{tpLang,kama,tp3,tp4}
In~\cite{interview}, Sonja Lang (the creator of TP),
describes that she has seen the language been used successfully
in the context of management, creation of texts, legal texts,
etc.

@@

There are standard guidelines for pronunciation,
but the language allows for considerable allophonic
variation.
For example, /p t k s l/ might be pronounced
% [p t k s l] or [b d g z !R].
[p t k s l] or [b d g z \textipa{\!R}].% ſ].
Especially for poetry, one might consider
j and w to be vowels
(e.g. j as 'i' and w as 'u').

@@

\section{Conclusions and further work}\label{conc}
This document presented a potentially novel description of Toki Pona (TP)
language in Section~\ref{basics}, and innovative and useful software routines for dealing with
the language in Section~\ref{hacks}.
As a minimal conlang, TP is very convenient for cognitive
experiments
through linguistic semantics and
for devising tools for analysis, creation and visualization,
a fact to which this document is evidence.
Other conlangs, and even non-planned languages might benefit from
TP and this content in many ways, e.g. one might
synthesize TP texts and translate them as convenient,
adapt the routines to obtain statistics of the vocabulary,
or take advantage of the language description to devise new conlangs
or stylistic guidelines for any language.

At the moment, the framework in~\cite{prv} is being used, by the author, for TP chatter bots
and for semantic explorations of domain knowledge.

@@

  \item Explore a minimal conlang to describe software

@@

% Bibliography
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-bibliography}
